Escape Analysis and Specialization in a Tracing JIT

### 2010-09-27 talk/pepm2011/paper.tex
refactor a number of things, add many XXXs

@@

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. The PyPy
project is an environment for implementing dynamic programming languages. It's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient VM that also contains a just-in-time compiler. This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}.

@@

\begin{enumerate}
    \item An efficient and effective algorithm for removing objects allocations in a tracing JIT.
    \item XXX
\end{enumerate}

@@

@@

The loop iterates \texttt{y} times, and computes something in the process. To
understand the reason why executing this function is slow, here is the trace
that is produced by the tracing JIT when executing the function with \texttt{y}
being a \texttt{BoxedInteger}: XXX make it clear that this is really a trace specific for BoxedInteger

@@

    # inside f: y.is_positive()
    guard_class(p15, BoxedInteger)
        # inside BoxedInteger.is_positive
        i16 = getfield_gc(p15, intval)
        i17 = int_gt(i16, 0)
    # inside f
    guard_true(i17)
    jump(p15, p10)
\end{verbatim}
\caption{unoptimized trace for the simple object model}
\end{figure}

(indentation corresponds to the stack level of the traced functions).

The trace is inefficient for a couple of reasons. One problem is that it checks
repeatedly and redundantly for the class of the objects around, using a
\texttt{guard\_class} instruction. In addition, some new \texttt{BoxedInteger} instances are
constructed using the \texttt{new} operation, only to be used once and then forgotten
a bit later. In the next section, we will see how this can be improved upon,
using escape analysis.

@@

\begin{figure}
\includegraphics{figures/obj-lifetime.pdf}

@@

The figure shows a trace before optimization, together with the lifetime of
various kinds of objects created in the trace. It is executed from top to
bottom. At the bottom, a jump is used to execute the same loop another time.
For clarity, the figure shows two iterations of the loop.
The loop is executed until one of the guards in the trace fails, and the
execution is aborted.

Some of the operations within this trace are \texttt{new} operations, which each create a
new instance of some class. These instances are used for a while, e.g. by
calling methods on them, reading and writing their fields. Some of these
instances escape, which means that they are stored in some globally accessible
place or are passed into a function.

Together with the \texttt{new} operations, the figure shows the lifetimes of the
created objects. Objects in category 1 live for a while, and are then just not
used any more. The creation of these objects is removed by the
optimization described in the last section.

Objects in category 2 live for a while and then escape. The optimization of the
last section deals with them too: the \texttt{new} that creates them and
the field accesses are deferred, until the point where the object escapes.

The objects in category 3 and 4 are in principle like the objects in category 1
and 2. They are created, live for a while, but are then passed as an argument
to the \texttt{jump} operation. In the next iteration they can either die (category
3) or escape (category 4).

@@

The main insight to improve the code shown in the last section is that some of
the objects created in the trace using a \texttt{new} operation don't survive very
long and are collected by the garbage collector soon after their allocation.
Moreover, they are used only inside the loop, thus we can easily prove that
nobody else in the program stores a reference to them. The
idea for improving the code is thus to analyze which objects never escape the
loop and may thus not be allocated at all.

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a virtual object is constructed. The virtual object summarizes the
shape of the object that is allocated at this position in the original trace,
and is used by the escape analysis to improve the trace. The shape describes
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
object, that shape summary is thus updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
from the virtual object, and the operation is also removed.

@@

\begin{verbatim}
p15 = new(BoxedInteger)
setfield_gc(p15, i14, intval)
p10 = new(BoxedInteger)
setfield_gc(p10, i9, intval)
jump(p15, p10)
\end{verbatim}

Note how the operations for creating these two instances has been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the forced virtual objects have been removed (some \texttt{getfield\_gc} operations
and \texttt{guard\_class} operations).

The final optimized trace of the example looks like this:

\begin{verbatim}
# arguments to the trace: p0, p1
guard_class(p1, BoxedInteger)
i2 = getfield_gc(p1, intval)
guard_class(p0, BoxedInteger)
i3 = getfield_gc(p0, intval)
i4 = int_add(i2, i3)
i9 = int_add(i4, -100)

guard_class(p0, BoxedInteger)
i12 = getfield_gc(p0, intval)
i14 = int_add(i12, -1)

i17 = int_gt(i14, 0)
guard_true(i17)
p15 = new(BoxedInteger)
setfield_gc(p15, i14, intval)
p10 = new(BoxedInteger)
setfield_gc(p10, i9, intval)
jump(p15, p10)
\end{verbatim}

@@

@@

@@

\begin{figure}
\includegraphics{figures/step1.pdf}
\end{figure}

>>>>

XXX define "escape analysis"

@@

XXX need to re-target introduction a bit to fit PEPMs focus

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. This
approach is also the one taken by the PyPy project, which is an environment for
implementing dynamic programming languages. PyPy's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient VM that also contains a just-in-time compiler. This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}.

@@

\begin{enumerate}
    \item An efficient and effective algorithm for removing object allocations in a tracing JIT.
    \item XXX
\end{enumerate}

@@

bridges?

@@

The loop iterates \texttt{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \texttt{is\_positive} and even two for each
call to \texttt{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \texttt{f} is simply a number of
float or integer additions.

@@

    # inside f: y.is_positive()
    guard_class(p15, BoxedInteger)
        # inside BoxedInteger.is_positive
        i16 = getfield_gc(p15, intval)
        i17 = int_gt(i16, 0)
    # inside f
    guard_true(i17)
    jump(p15, p10)
\end{verbatim}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new}, \texttt{set/getfield\_gc} and
\texttt{guard\_class} operations.

Note how the functions that are called by \texttt{f} are automatically inlined
into the trace. The method calls are always preceded by a \texttt{guard\_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\texttt{guard\_class} performs a precise
class check, not checking for subclasses} These guards make the trace specific
to the situation where \texttt{y} is really a \texttt{BoxedInteger}, it can
already be said to be specialized for \texttt{BoxedIntegers}. When the trace is
turned into machine code and then executed with \texttt{BoxedFloats}, the
first \texttt{guard\_class} instruction will fail and execution will continue
using the interpreter.

XXX simplify traces a bit more
get rid of \_gc suffix in set/getfield\_gc

In the next section, we will see how this can be improved upon, using escape
analysis. XXX

@@

\begin{figure}
\includegraphics[scale=0.7]{figures/obj-lifetime.pdf}

@@

Figure~\ref{fig:lifetimes} shows a trace before optimization, together with the
lifetime of various kinds of objects created in the trace. It is executed from
top to bottom. At the bottom, a jump is used to execute the same loop another
time (for clarity, the figure shows two iterations of the loop). The loop is
executed until one of the guards in the trace fails, and the execution is
aborted and interpretation resumes.

Some of the operations within this trace are \texttt{new} operations, which each
create a new instance of some class. These instances are used for a while, e.g.
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a function.

Together with the \texttt{new} operations, the figure shows the lifetimes of the
created objects. The objects that are created within a trace using \texttt{new}
fall into one of several categories:

\begin{itemize}
    \item Category 1: Objects that live for a while, and are then just not
    used any more.

    \item Category 2: Objects that live for a while and then escape.

    \item Category 3: Objects that live for a while, survive across the jump to
    the beginning of the loop, and are then not used any more.

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used\footnote{In theory, the
    approach of Section~\ref{sec:XXX} works also for objects that live for
    exactly $n>1$ iterations and then don't escape, but we expect this to be a
    very rare case, so we do not handle it.}
\end{itemize}

The objects that are allocated in the example trace in
Figure~\ref{fig:unopt-trace} fall into categories 1 and 3. Objects stored in
\texttt{p5, p6, p11 XXX} are in category 1, objects in \texttt{p10, p15} are in
category 3.

The creation of objects in category 1 is removed by the optimization described
in Section~\ref{sec:virtuals}. XXX

@@

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long and are collected by the garbage collector
soon after their allocation. Moreover, they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is thus to analyze which objects fall in category 1 and which may thus
not be allocated at all.

XXX is "symbolic execution" the right word to drop?

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a virtual object is constructed. The virtual object summarizes the
shape of the object that is allocated at this position in the original trace,
and is used by the optimization to improve the trace. The shapes describe
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
object, that shape summary is thus updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
from the virtual object, and the operation is also removed.

XXX what happens on a guard\_class?

@@

\texttt{
\begin{tabular}{l} 
$p_{15}$ = new(BoxedInteger) \\
setfield\_gc($p_{15}$, $i_{14}$, intval) \\
$p_{10}$ = new(BoxedInteger) \\
setfield\_gc($p_{10}$, $i_{9}$, intval) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the forced virtual objects have been removed (some \texttt{getfield\_gc} operations
and \texttt{guard\_class} operations).

\begin{figure}
\includegraphics{figures/step1.pdf}
\label{fig:step1}
\caption{Resulting Trace After Allocation Removal}
\end{figure}

The final optimized trace of the example can be seen in
Figure~\ref{fig:step1}.

@@

\subsection{Algorithm}
\label{sub:Algorithm}

XXX want some sort of pseudo-code

% subsection Algorithm (end)

@@

XXX Category 2 The optimization of
Section~\ref{sec:virtuals} deals with them too: the \texttt{new} that creates them and
the field accesses are deferred, until the point where the object escapes.

@@
